---
title: "Differential Calculus"
subtitle: "Residuals, loss functions, derivatives, and optimization"
---


## Activities

### Activity 1: Why residuals need a loss function
Students fit a line, compute residuals, and observe that positive and negative errors can cancel. They compare absolute error and squared error.

### Activity 2: Smooth versus nonsmooth loss
Students graph absolute value loss and squared loss. They discuss why differentiability matters for optimization algorithms.

### Activity 3: Find the best slope
Students treat slope as a variable and plot mean squared error over many possible slopes. They use the graph to locate a minimizer before connecting the idea to derivatives.

## Ethics and equity prompt

A small average error can hide systematic errors for subgroups. Ask students to compare residuals by species or another group variable and discuss fairness in predictive performance.

## R code

```{r}
source("../R/setup.R")
fish <- load_fish()
fit <- lm(Weight ~ Length1, data = fish)
aug <- broom::augment(fit, data = fish)
aug |> summarise(sum_residuals = sum(.resid), mean_abs_error = mean(abs(.resid)), mse = mean(.resid^2)) |>
  knitr::kable(digits = 3)
```

```{r}
tibble(x = seq(-10, 10, length.out = 401)) |>
  pivot_longer(cols = x, names_to = "variable", values_to = "error") |>
  mutate(`absolute loss` = abs(error), `squared loss` = error^2) |>
  pivot_longer(c(`absolute loss`, `squared loss`), names_to = "loss", values_to = "value") |>
  ggplot(aes(error, value)) + geom_line() + facet_wrap(~loss, scales = "free_y") +
  labs(title = "Two common loss functions", x = "Prediction error", y = "Loss")
```

```{r}
slopes <- tibble(slope = seq(0, 40, length.out = 250)) |>
  mutate(intercept = mean(fish$Weight) - slope * mean(fish$Length1),
         mse = map2_dbl(slope, intercept, ~mean((fish$Weight - (.y + .x * fish$Length1))^2)))
slopes |> ggplot(aes(slope, mse)) + geom_line() +
  labs(title = "Mean squared error as a function of slope", y = "MSE")
slopes |> slice_min(mse, n = 1) |> knitr::kable(digits = 3)
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
