---
title: "Integral Calculus"
subtitle: "Accumulation, numerical integration, convergence, and gradient descent"
---


## Activities

### Activity 1: Area as accumulated prediction error
Students interpret the area under an absolute residual curve as accumulated model error over a range of inputs.

### Activity 2: Numerical integration and approximation error
Students compare left sums, trapezoidal rule, and Simpson-style thinking for approximating an area under a curve.

### Activity 3: Convergence in simple gradient descent
Students iterate a simple optimization rule and connect convergence to sequences and limiting behavior.

## Ethics and equity prompt

Approximation choices can shape conclusions. Ask students when a coarse approximation is acceptable and when it may obscure harm or uncertainty.

## R code

```{r}
source("../R/setup.R")
f <- function(x) (x - 3)^2 + 2
xgrid <- tibble(x = seq(0, 6, length.out = 200), y = f(x))
xgrid |> ggplot(aes(x, y)) + geom_line() +
  labs(title = "Function to approximate with numerical integration")
```

```{r}
trap_area <- function(a, b, n) {
  x <- seq(a, b, length.out = n + 1)
  y <- f(x)
  sum((y[-1] + y[-length(y)]) / 2 * diff(x))
}
tibble(n = c(4, 8, 16, 32, 64, 128)) |>
  mutate(trapezoid_area = map_dbl(n, ~trap_area(0, 6, .x))) |>
  knitr::kable(digits = 5)
```

```{r}
gradient_descent <- function(start = 8, rate = .1, steps = 30) {
  x <- numeric(steps); x[1] <- start
  for (i in 2:steps) x[i] <- x[i-1] - rate * 2 * (x[i-1] - 3)
  tibble(step = 1:steps, x = x, objective = f(x))
}
gradient_descent() |> ggplot(aes(step, objective)) + geom_line() + geom_point() +
  labs(title = "Convergence of a simple gradient descent sequence")
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
