---
title: "Introductory Programming"
subtitle: "Data types, functions, iteration, debugging, reproducibility, and AI tools"
---


## Activities

### Activity 1: Data as objects
Students inspect vectors, data frames, factors, and missing values. They connect programming data types to statistical meaning.

### Activity 2: Write a prediction function
Students write a small R function that predicts fish weight from length using a fitted model.

### Activity 3: Debugging and reproducibility
Students intentionally break code, read error messages, and practice using comments, seeds, file paths, and Quarto chunks.

## Ethics and equity prompt

Students compare human-written and AI-assisted code. They ask: what needs verification, what assumptions are hidden, and how should credit or disclosure be handled?

## R code

```{r}
source("../R/setup.R")
fish <- load_fish()
str(fish)
```

```{r}
fit <- lm(Weight ~ Length1, data = fish)
predict_weight <- function(length_cm) {
  predict(fit, newdata = tibble(Length1 = length_cm))
}
predict_weight(c(10, 25, 40))
```

```{r}
# A reproducible mini-analysis pipeline
analyze_species <- function(species_name) {
  fish |> filter(Species == species_name) |>
    summarise(n = n(), mean_weight = mean(Weight), sd_weight = sd(Weight))
}
map_dfr(unique(fish$Species), analyze_species, .id = "analysis_id") |> knitr::kable(digits = 2)
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
