---
title: "Introductory Statistics"
subtitle: "Visualization, cleaning, inference, regression, and prediction"
---


## Activities

### Activity 1: Cleaning changes conclusions
Students intentionally introduce and then correct data issues such as impossible values, miscoded categories, and missing values. They compare summaries before and after cleaning.

### Activity 2: Regression for prediction
Students fit and interpret a regression model, distinguish association from causation, and evaluate prediction error.

### Activity 3: Sampling variability simulation
Students repeatedly sample from the dataset and visualize the distribution of sample means or slopes.

## Ethics and equity prompt

Data cleaning is not neutral. Ask students to document every cleaning decision and explain how each decision could affect conclusions.

## R code

```{r}
source("../R/setup.R")
fish <- load_fish()
fish_dirty <- fish |> mutate(Weight = if_else(row_number() == 1, -99, Weight), Species = if_else(row_number() == 2, "perch", Species))
fish_dirty |> summarise(min_weight = min(Weight), species_count = n_distinct(Species)) |> knitr::kable()
fish_clean <- fish_dirty |> mutate(Weight = na_if(Weight, -99), Species = str_to_title(Species))
fish_clean |> summarise(min_weight = min(Weight, na.rm = TRUE), species_count = n_distinct(Species)) |> knitr::kable()
```

```{r}
fit <- lm(Weight ~ Length1 + Height + Width, data = fish_clean)
broom::tidy(fit) |> knitr::kable(digits = 3)
```

```{r}
set.seed(2026)
boot_slopes <- replicate(500, {
  samp <- fish_clean |> slice_sample(n = nrow(fish_clean), replace = TRUE)
  coef(lm(Weight ~ Length1, data = samp))[2]
})
tibble(slope = boot_slopes) |> ggplot(aes(slope)) + geom_histogram(bins = 30) +
  labs(title = "Bootstrap distribution of regression slopes")
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
