---
title: "Linear Algebra"
subtitle: "Vectors, dot products, cosine similarity, clustering, PCA/SVD"
---


## Activities

### Activity 1: Dot product as similarity
Students compute dot products and cosine similarity for fish measurements, then interpret similarity as angle rather than raw magnitude.

### Activity 2: K-means clustering
Students cluster observations using length, height, and width. They compare clusters to species labels and discuss why unsupervised learning may not recover human-defined categories.

### Activity 3: PCA as change of basis
Students use principal components to represent high-dimensional measurements in two dimensions.

## Ethics and equity prompt

Similarity measures are design choices. Ask students how scaling, variable selection, and missing context can change who is considered similar to whom.

## R code

```{r}
source("../R/setup.R")
fish <- load_fish()
X <- fish |> select(Length1, Length2, Length3, Height, Width) |> scale()
cosine <- function(x, y) sum(x*y) / sqrt(sum(x^2) * sum(y^2))
cosine(X[1,], X[2,])
```

```{r}
set.seed(2026)
km <- kmeans(X, centers = 3, nstart = 25)
fish |> mutate(cluster = factor(km$cluster)) |>
  count(Species, cluster) |>
  pivot_wider(names_from = cluster, values_from = n, values_fill = 0) |>
  knitr::kable()
```

```{r}
pca <- prcomp(X)
augment(pca, data = fish) |>
  ggplot(aes(.fittedPC1, .fittedPC2, color = Species)) + geom_point() +
  labs(title = "PCA of fish measurements", x = "PC1", y = "PC2")
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
