---
title: "Multivariable Calculus"
subtitle: "Partial derivatives, gradients, contours, and gradient descent"
---


## Activities

### Activity 1: Contour maps of loss landscapes
Students visualize a two-parameter loss surface and identify local/global minima.

### Activity 2: Follow the negative gradient
Students compute or approximate gradients and trace the path of gradient descent on a contour plot.

### Activity 3: Constraints and model tradeoffs
Students discuss constrained optimization as a model design problem: improve accuracy while limiting cost, complexity, or inequitable impact.

## Ethics and equity prompt

Optimization requires an objective. Ask: what did we choose to optimize, what did we leave out, and who might be affected by that choice?

## R code

```{r}
loss <- function(a, b) (a - 2)^2 + 2 * (b + 1)^2 + .5 * a * b
surface <- expand_grid(a = seq(-4, 6, length.out = 120), b = seq(-5, 3, length.out = 120)) |>
  mutate(loss = loss(a, b))
surface |> ggplot(aes(a, b, z = loss)) + geom_contour_filled(bins = 15) +
  labs(title = "Two-parameter loss landscape", x = "Parameter a", y = "Parameter b")
```

```{r}
step <- function(a, b, rate = .12) {
  da <- 2 * (a - 2) + .5 * b
  db <- 4 * (b + 1) + .5 * a
  c(a - rate * da, b - rate * db)
}
path <- tibble(iter = 1:30, a = NA_real_, b = NA_real_)
path$a[1] <- -3; path$b[1] <- 2.5
for (i in 2:nrow(path)) {
  nxt <- step(path$a[i-1], path$b[i-1])
  path$a[i] <- nxt[1]; path$b[i] <- nxt[2]
}
surface |> ggplot(aes(a, b, z = loss)) + geom_contour() +
  geom_path(data = path, aes(a, b), inherit.aes = FALSE) + geom_point(data = path, aes(a, b), inherit.aes = FALSE) +
  labs(title = "Gradient descent path")
```


## Assessment ideas

- Instructor pre/post: explain the data science concept in course language and identify one classroom integration point.
- Perceived value: rate feasibility, relevance, and equity-mindedness of the activity.
- Classroom artifact: submit an adapted project, discussion prompt, quiz item, or notebook cell.
- Reflection: describe one modification for students with varied coding backgrounds.
